\texttt{MSnbase}, efficient R-based access and manipulation of raw mass spectrometry data

### 2020-04-24 msnbase2.Rnw
proceed with on-disk benchmarking section

\documentclass[journal=jacsat,manuscript=article]{achemso}

@@

\author{Laurent Gatto}
\email{laurent.gatto@uclouvain.be}
\affiliation[UCLouvain]{de Duve Institute, Universit\'e catholique de Louvain, Brussels, Belgium}
\author{Sebastian Gibb}
\affiliation[University of Greifswald]{Department of Anaesthesiology and Intensive Care of the University Medicine Greifswald, Germany}
\author{Johannes Rainer}
\affiliation[EURAC]{Institute for Biomedicine, Eurac Research, Affiliated Institute of the University of L\"ubeck, Bolzano, Italy}

@@

\begin{abstract} %% 200 words max
  We present version 2 of the \texttt{MSnbase} R/Bioconductor
  package. \texttt{MSnbase} provides infrastructure for the
  manipulation, processing and visualisation of mass spectrometry
  data. Here we present how the new \textit{on disk} infrastructure
  allows the handling of hundreds on commodity hardware and present
  some application of the package.
\end{abstract}

@@

\subsection{On disk backend}

Efficient low level access: in memory vs on disk mode (using mzR),
benchmarking, used in ms-based proteomics and metabolomics.

\subsection{Use cases}

Example applications:

- visualisation
- large scale data analyses (metabolomics, Johannes)
- boxcar prototyping

@@

The version of \texttt{MSnbase} used in this manuscritp is version
\Sexpr{packageVersion("MSnbase")}. The main features presented here
were available since version 2.0.

>>>>

\documentclass[journal=jacsat,manuscript=article]{achemso}\usepackage[]{graphicx}\usepackage[]{color}
%% maxwidth is the original width if it is less than linewidth
%% otherwise use linewidth (to make sure the graphics do not exceed the margin)
\makeatletter
\def\maxwidth{ %
  \ifdim\Gin@nat@width>\linewidth
    \linewidth
  \else
    \Gin@nat@width
  \fi
}
\makeatother

%% \definecolor{fgcolor}{rgb}{0.345, 0.345, 0.345}
%% \newcommand{\hlnum}[1]{\textcolor[rgb]{0.686,0.059,0.569}{#1}}%
%% \newcommand{\hlstr}[1]{\textcolor[rgb]{0.192,0.494,0.8}{#1}}%
%% \newcommand{\hlcom}[1]{\textcolor[rgb]{0.678,0.584,0.686}{\textit{#1}}}%
%% \newcommand{\hlopt}[1]{\textcolor[rgb]{0,0,0}{#1}}%
%% \newcommand{\hlstd}[1]{\textcolor[rgb]{0.345,0.345,0.345}{#1}}%
%% \newcommand{\hlkwa}[1]{\textcolor[rgb]{0.161,0.373,0.58}{\textbf{#1}}}%
%% \newcommand{\hlkwb}[1]{\textcolor[rgb]{0.69,0.353,0.396}{#1}}%
%% \newcommand{\hlkwc}[1]{\textcolor[rgb]{0.333,0.667,0.333}{#1}}%
%% \newcommand{\hlkwd}[1]{\textcolor[rgb]{0.737,0.353,0.396}{\textbf{#1}}}%
%% \let\hlipl\hlkwb

%% \usepackage{framed}
%% \makeatletter
%% \newenvironment{kframe}{%
%%  \def\at@end@of@kframe{}%
%%  \ifinner\ifhmode%
%%   \def\at@end@of@kframe{\end{minipage}}%
%%   \begin{minipage}{\columnwidth}%
%%  \fi\fi%
%%  \def\FrameCommand##1{\hskip\@totalleftmargin \hskip-\fboxsep
%%  \colorbox{shadecolor}{##1}\hskip-\fboxsep
%%      % There is no \\@totalrightmargin, so:
%%      \hskip-\linewidth \hskip-\@totalleftmargin \hskip\columnwidth}%
%%  \MakeFramed {\advance\hsize-\width
%%    \@totalleftmargin\z@ \linewidth\hsize
%%    \@setminipage}}%
%%  {\par\unskip\endMakeFramed%
%%  \at@end@of@kframe}
%% \makeatother

%% \usepackage{xcolor}

%% \definecolor{shadecolor}{rgb}{.97, .97, .97}
%% \definecolor{messagecolor}{rgb}{0, 0, 0}
%% \definecolor{warningcolor}{rgb}{1, 0, 1}
%% \definecolor{errorcolor}{rgb}{1, 0, 0}
%% \newenvironment{knitrout}{}{} % an empty environment to be redefined in TeX

\usepackage{subfig}
\usepackage{graphicx}
\usepackage{alltt}

@@

\author{Laurent Gatto}
\email{laurent.gatto@uclouvain.be}
\affiliation[UCLouvain]{Computational Biology Unit, de Duve Institute, Universit\'e catholique de Louvain, Brussels, Belgium}
\author{Sebastian Gibb}
\affiliation[University of Greifswald]{Department of Anaesthesiology and Intensive Care of the University Medicine Greifswald, Germany}
\author{Johannes Rainer}
\affiliation[EURAC]{Institute for Biomedicine, Eurac Research, Affiliated Institute of the University of L\"ubeck, Bolzano, Italy}

@@

\begin{abstract} %% 200 words max
  We present version 2 of the \texttt{MSnbase} R/Bioconductor
  package. \texttt{MSnbase} provides infrastructure for the
  manipulation, processing and visualisation of mass spectrometry
  data. Here we present how the new \textit{on-disk} infrastructure
  allows the handling of hundreds on commodity hardware and present
  some application of the package.
\end{abstract}

<<setup, include = FALSE, message = FALSE>>=
knitr::opts_chunk$set(echo = FALSE, cache = FALSE)
library("xtable")
library("microbenchmark")
library("tidyverse")
@

@@

\subsection{On-disk backend}

The main feature in version 2 of the \texttt{MSnbase} package was the
addition of different backends for raw data storage, namely
\textit{in-memory} and on-disk. The following code chunk
demonstrates how to create two \texttt{MSnExp} objects, tailored to
manage mass spectrometry experiments, storing data in-memory or
on-disk.

<<readMSData, eval = FALSE, echo = TRUE>>=
library("MSnbase")
raw_mem <- readMSData("file.mzML", mode = "inMemory")
raw_dsk <- readMSData("file.mzML", mode = "onDisk")
@

The former is the legacy storage mode, implemented in the first
version of the package, that loads all the raw data and the metadata
in memory. This solution doesn't scale for modern large dataset, and
was complemented by the on-disk backend, that only load
metadata in memory and accesses the spectra in the original files when
needed. There are two direct benefits using the on-disk
backend. Table \ref{tab:bench1} shows 4-fold faster reading times
(left) and 10-fold smaller memory footprint (right).

<<time_sz>>=
load("bench_time_sz.rda")
time <- sapply(time_sz, function(x) x[, "time"])
sz <- sapply(time_sz, function(x) x[, "sz"])
colnames(sz) <- colnames(time) <- c(1, 5, 10)
rownames(sz) <- rownames(time) <- c("in-memory", "on-disk")
sz <- round(sz/(1014^2), 2)
print(xtable(time), floating = FALSE, file = "tabs/t_read.tex")
print(xtable(sz), floating = FALSE, include.rownames = FALSE, file = "tabs/sz.tex")
@

\begin{table}[ht]
\centering
\subfloat[Reading time (s)]{\label{tab:tread}\scalebox{1}{\input{./tabs/t_read}}}\quad
\subfloat[Object size (MB)]{\label{tab:sz}\scalebox{1}{\input{./tabs/sz}}}
\caption{Reading time (in seconds) (left) and data size in memory (in
  MB) to read/store 1, 5 and 10 files containing 1431 MS1 (on-disk
  only) and 6103 MS2 (on-disk and in-memory) spectra. }
\label{tab:bench1}
\end{table}

The on-disk backend also offers efficient data manipulation by way of
\textit{lazy processing}. Operations on the raw data are stored in a
processing queue and only effectively applied when raw data is
accessed on disk. As an example, the following short analysis
pipeline, that can equally be applied to on in-memory or on-disk data
retains MS2 spectra acquired between 1000 and 3000 seconds, extract
the M/Z range corresponding to the TMT 6-plex range and focuses on the
MS2 spectra with a precursor intensity greater than $11 \times 10^6$
(the median precursor intensity).

<<filter, eval = FALSE, echo = TRUE>>=
x <- x_dsk %>%
    filterRt(c(1000, 3000)) %>%
    filterMz(120, 135)
x[precursorIntensity(x) > 11e6, ]
@

As shown on figure \ref{fig:filt}, this lazy mechanism si
significantly faster than its application on in-memory data.

\begin{figure}[ht]
  \centering
<<plot_filter, message = FALSE, out.width = '50%', fig.width = 5, fig.height = 3>>=
load("bench_t_filt.rda")
ggplot2::autoplot(t_filt)
@
\caption{Filtering benchmark assessed over 10 interations on in-memory (top) and on-disk (bottom) data containing 6103 MS2 spectra.}
\label{fig:filt}
\end{figure}

The advantageous reading and execution times and memory footprint of
the on-disk backend are possible by avoiding unnecessary access to the
raw data. Once access to the spectra M/Z and intensity values become
mandatory (for example for plotting), then the in-memory backend
becomes more efficient, as illustrated on figure
\ref{fig:access}. This gain is maximal when the whole dataset is the
be accessed (i.e. all spectra are already in memory) and negligeable
when large fractions of the data need to be subset.

\begin{figure}[ht]
  \centering
<<plot_access, message = FALSE, out.width = '75%', fig.width = 5, fig.height = 3>>=
load("bench_t_access.rda")
tibble(time = microbenchmark:::convert_to_unit(t_access$time, "ms"),
       expr = as.character(t_access$expr)) %>%
    mutate(n = sub("^access_.+_", "", expr)) %>%
    mutate(n = sub("all", 6103, n)) %>%
    mutate(mode = if_else(grepl("mem", expr), "in-memory", "on-disk")) %>%
    ggplot(aes(x = n, y = time, fill = log10(as.numeric(n)))) + 
    ggplot2::geom_violin() + 
    ggplot2::scale_y_log10() +
    ylab("Time [millisec]") +
    xlab("Number of spectra (out of 6103)") +
    facet_wrap(~ mode) +
    theme(legend.position = "none")
@
\caption{Access time to spectra for the in-memory (left) and on-disk
  (right) backends for 1, 10, 100 1000, 5000 and all 6103 spectra.}
\label{fig:access}
\end{figure}

Enables large scale data analyses (metabolomics example, see Johannes)

\subsection{Prototyping}

See https://github.com/lgatto/msnbase\_boxcar.

\subsection{Visualisation}

Examples: 3D MSmap, boxcar, centroiding vignette

@@

The version of \texttt{MSnbase} used in this manuscritp is version
2.10.0. The main features presented here
were available since version 2.0.

### 2020-04-24 msnbase2.tex
add benchmarking scipts and results
proceed with on-disk benchmarking section

\documentclass[journal=jacsat,manuscript=article]{achemso}\usepackage[]{graphicx}\usepackage[]{color}
%% maxwidth is the original width if it is less than linewidth
%% otherwise use linewidth (to make sure the graphics do not exceed the margin)
\makeatletter
\def\maxwidth{ %
  \ifdim\Gin@nat@width>\linewidth
    \linewidth
  \else
    \Gin@nat@width
  \fi
}
\makeatother

@@

\usepackage{alltt}

@@

\abbreviations{}
\keywords{Bioconductor, mass spectrometry, software, reproducible research} %% up to 10 keywords
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
\begin{document}

@@

\begin{abstract} %% 200 words max
  We present version 2 of the \texttt{MSnbase} R/Bioconductor
  package. \texttt{MSnbase} provides infrastructure for the
  manipulation, processing and visualisation of mass spectrometry
  data. Here we present how the new \textit{on disk} infrastructure
  allows the handling of hundreds on commodity hardware and present
  some application of the package.
\end{abstract}

@@

\subsection{On disk backend}

Efficient low level access: in memory vs on disk mode (using mzR),
benchmarking, used in ms-based proteomics and metabolomics.

\subsection{Use cases}

Example applications:

- visualisation
- large scale data analyses (metabolomics, Johannes)
- boxcar prototyping

>>>>

\documentclass[journal=jacsat,manuscript=article]{achemso}\usepackage[]{graphicx}\usepackage[]{color}
% maxwidth is the original width if it is less than linewidth
% otherwise use linewidth (to make sure the graphics do not exceed the margin)
\makeatletter
\def\maxwidth{ %
  \ifdim\Gin@nat@width>\linewidth
    \linewidth
  \else
    \Gin@nat@width
  \fi
}
\makeatother

@@

\usepackage{alltt}\usepackage[]{graphicx}\usepackage[]{color}
%% maxwidth is the original width if it is less than linewidth
%% otherwise use linewidth (to make sure the graphics do not exceed the margin)
\makeatletter
\def\maxwidth{ %
  \ifdim\Gin@nat@width>\linewidth
    \linewidth
  \else
    \Gin@nat@width
  \fi
}
\makeatother

%% \definecolor{fgcolor}{rgb}{0.345, 0.345, 0.345}
%% \newcommand{\hlnum}[1]{\textcolor[rgb]{0.686,0.059,0.569}{#1}}%
%% \newcommand{\hlstr}[1]{\textcolor[rgb]{0.192,0.494,0.8}{#1}}%
%% \newcommand{\hlcom}[1]{\textcolor[rgb]{0.678,0.584,0.686}{\textit{#1}}}%
%% \newcommand{\hlopt}[1]{\textcolor[rgb]{0,0,0}{#1}}%
%% \newcommand{\hlstd}[1]{\textcolor[rgb]{0.345,0.345,0.345}{#1}}%
%% \newcommand{\hlkwa}[1]{\textcolor[rgb]{0.161,0.373,0.58}{\textbf{#1}}}%
%% \newcommand{\hlkwb}[1]{\textcolor[rgb]{0.69,0.353,0.396}{#1}}%
%% \newcommand{\hlkwc}[1]{\textcolor[rgb]{0.333,0.667,0.333}{#1}}%
%% \newcommand{\hlkwd}[1]{\textcolor[rgb]{0.737,0.353,0.396}{\textbf{#1}}}%
%% \let\hlipl\hlkwb

%% \usepackage{framed}
%% \makeatletter
%% \newenvironment{kframe}{%
%%  \def\at@end@of@kframe{}%
%%  \ifinner\ifhmode%
%%   \def\at@end@of@kframe{\end{minipage}}%
%%   \begin{minipage}{\columnwidth}%
%%  \fi\fi%
%%  \def\FrameCommand##1{\hskip\@totalleftmargin \hskip-\fboxsep
%%  \colorbox{shadecolor}{##1}\hskip-\fboxsep
%%      % There is no \\@totalrightmargin, so:
%%      \hskip-\linewidth \hskip-\@totalleftmargin \hskip\columnwidth}%
%%  \MakeFramed {\advance\hsize-\width
%%    \@totalleftmargin\z@ \linewidth\hsize
%%    \@setminipage}}%
%%  {\par\unskip\endMakeFramed%
%%  \at@end@of@kframe}
%% \makeatother

%% \usepackage{xcolor}

%% \definecolor{shadecolor}{rgb}{.97, .97, .97}
%% \definecolor{messagecolor}{rgb}{0, 0, 0}
%% \definecolor{warningcolor}{rgb}{1, 0, 1}
%% \definecolor{errorcolor}{rgb}{1, 0, 0}
%% \newenvironment{knitrout}{}{} % an empty environment to be redefined in TeX

\usepackage{subfig}
\usepackage{graphicx}
\usepackage{alltt}

@@

\abbreviations{}
\keywords{Bioconductor, mass spectrometry, software, reproducible research} %% up to 10 keywords
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
\begin{document}

@@

\begin{abstract} %% 200 words max
  We present version 2 of the \texttt{MSnbase} R/Bioconductor
  package. \texttt{MSnbase} provides infrastructure for the
  manipulation, processing and visualisation of mass spectrometry
  data. Here we present how the new \textit{on-disk} infrastructure
  allows the handling of hundreds on commodity hardware and present
  some application of the package.
\end{abstract}

@@

\subsection{On-disk backend}

The main feature in version 2 of the \texttt{MSnbase} package was the
addition of different backends for raw data storage, namely
\textit{in-memory} and on-disk. The following code chunk
demonstrates how to create two \texttt{MSnExp} objects, tailored to
manage mass spectrometry experiments, storing data in-memory or
on-disk.

\begin{knitrout}
\definecolor{shadecolor}{rgb}{0.969, 0.969, 0.969}\color{fgcolor}\begin{kframe}
\begin{alltt}
\hlkwd{library}\hlstd{(}\hlstr{"MSnbase"}\hlstd{)}
\hlstd{raw_mem} \hlkwb{<-} \hlkwd{readMSData}\hlstd{(}\hlstr{"file.mzML"}\hlstd{,} \hlkwc{mode} \hlstd{=} \hlstr{"inMemory"}\hlstd{)}
\hlstd{raw_dsk} \hlkwb{<-} \hlkwd{readMSData}\hlstd{(}\hlstr{"file.mzML"}\hlstd{,} \hlkwc{mode} \hlstd{=} \hlstr{"onDisk"}\hlstd{)}
\end{alltt}
\end{kframe}
\end{knitrout}

The former is the legacy storage mode, implemented in the first
version of the package, that loads all the raw data and the metadata
in memory. This solution doesn't scale for modern large dataset, and
was complemented by the on-disk backend, that only load
metadata in memory and accesses the spectra in the original files when
needed. There are two direct benefits using the on-disk
backend. Table \ref{tab:bench1} shows 4-fold faster reading times
(left) and 10-fold smaller memory footprint (right).

\begin{table}[ht]
\centering
\subfloat[Reading time (s)]{\label{tab:tread}\scalebox{1}{\input{./tabs/t_read}}}\quad
\subfloat[Object size (MB)]{\label{tab:sz}\scalebox{1}{\input{./tabs/sz}}}
\caption{Reading time (in seconds) (left) and data size in memory (in
  MB) to read/store 1, 5 and 10 files containing 1431 MS1 (on-disk
  only) and 6103 MS2 (on-disk and in-memory) spectra. }
\label{tab:bench1}
\end{table}

The on-disk backend also offers efficient data manipulation by way of
\textit{lazy processing}. Operations on the raw data are stored in a
processing queue and only effectively applied when raw data is
accessed on disk. As an example, the following short analysis
pipeline, that can equally be applied to on in-memory or on-disk data
retains MS2 spectra acquired between 1000 and 3000 seconds, extract
the M/Z range corresponding to the TMT 6-plex range and focuses on the
MS2 spectra with a precursor intensity greater than $11 \times 10^6$
(the median precursor intensity).

\begin{knitrout}
\definecolor{shadecolor}{rgb}{0.969, 0.969, 0.969}\color{fgcolor}\begin{kframe}
\begin{alltt}
\hlstd{x} \hlkwb{<-} \hlstd{x_dsk} \hlopt{%>%}
    \hlkwd{filterRt}\hlstd{(}\hlkwd{c}\hlstd{(}\hlnum{1000}\hlstd{,} \hlnum{3000}\hlstd{))} \hlopt{%>%}
    \hlkwd{filterMz}\hlstd{(}\hlnum{120}\hlstd{,} \hlnum{135}\hlstd{)}
\hlstd{x[}\hlkwd{precursorIntensity}\hlstd{(x)} \hlopt{>} \hlnum{11e6}\hlstd{, ]}
\end{alltt}
\end{kframe}
\end{knitrout}

As shown on figure \ref{fig:filt}, this lazy mechanism si
significantly faster than its application on in-memory data.

\begin{figure}[ht]
  \centering
\begin{knitrout}
\definecolor{shadecolor}{rgb}{0.969, 0.969, 0.969}\color{fgcolor}
\includegraphics[width=0.5\linewidth]{figure/plot_filter-1}

\end{knitrout}
\caption{Filtering benchmark assessed over 10 interations on in-memory (top) and on-disk (bottom) data containing 6103 MS2 spectra.}
\label{fig:filt}
\end{figure}

The advantageous reading and execution times and memory footprint of
the on-disk backend are possible by avoiding unnecessary access to the
raw data. Once access to the spectra M/Z and intensity values become
mandatory (for example for plotting), then the in-memory backend
becomes more efficient, as illustrated on figure
\ref{fig:access}. This gain is maximal when the whole dataset is the
be accessed (i.e. all spectra are already in memory) and negligeable
when large fractions of the data need to be subset.

\begin{figure}[ht]
  \centering
\begin{knitrout}
\definecolor{shadecolor}{rgb}{0.969, 0.969, 0.969}\color{fgcolor}
\includegraphics[width=0.75\linewidth]{figure/plot_access-1}

\end{knitrout}
\caption{Access time to spectra for the in-memory (left) and on-disk
  (right) backends for 1, 10, 100 1000, 5000 and all 6103 spectra.}
\label{fig:access}
\end{figure}

Enables large scale data analyses (metabolomics example, see Johannes)

\subsection{Prototyping}

See https://github.com/lgatto/msnbase\_boxcar.

\subsection{Visualisation}

Examples: 3D MSmap, boxcar, centroiding vignette
